Learning how to act: making good decisions with machine learning

### 2017-04-29 thesis/thesis_map.tex
bootcamp day2 update

\documentclass[11pt,a4paper]{book}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{dsfont}
\usepackage{listings}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{tabularx}
\usepackage{enumitem}
\usepackage{algorithm}% http://ctan.org/pkg/algorithm
\usepackage[noend]{algpseudocode}% http://ctan.org/pkg/algorithmicx
\usepackage{parskip}
\usepackage{tikz}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\usepackage{hyperref}
\usetikzlibrary{arrows,positioning} 
\usepackage{subcaption}

\pgfarrowsdeclarecombine{ring}{ring}{}{}{o}{o}

@@

@@

\renewcommand{\P}[1]{\operatorname{P}\left\{#1\right\}}

@@

\begin{document}
\def\ci{\perp\!\!\!\perp} % from Wikipedia
\newtheorem{theorem}{Theorem}[section]
\newtheorem{definition}{Definition}[section]
\maketitle

\section{This Thesis}
\begin{itemize}
\item clarify where causal inference fits within machine learing, for which problems it is required and ...
\end{itemize}

\section*{Uncategorized}
http://www.news.com.au/finance/money/costs/insurance-companies-secrets-spilt/news-story/f6ef17ae73e3a5636343845666532b91 insurance decisions voodoo because of lack of transparancy and absense of obvious causal link. Claim: People are more comfortable with desicion making on the basis of factors they believe to be causally relevet.

@@

\paragraph*{Motivating summary}

@@

\subsection*{Defining causality}
\begin{itemize}
\item widely debated in science and philosophy (FIND SOME REFERENCES)
\item what is explaination?
\item any model that aims to predict the outcome of an action or intervention in a system
\item I do not see the distinction between explaination and (causal) prediction. Explaination is all about the ability to compress and to generalize. The more a model can do this, the more we view as providing an understanding of the why. 
\item mediation?
\end{itemize}

@@

\begin{itemize}
\item Speech recognition (for systems like Siri or Google)
\item Machine translation 
\item Image classification
\item Forecasting the weather
\item Playing Go 
\item Identiying spam emails
\item Automated essay marking
\item Predicting the risk of death in patients with pnumonia.
\item Predicting who will re-offend on release from prison 
\item Predicting which customers will cease to be your customers
\item Demand prediction for inventory control
\item Predicting who will click on an ad
\item Financial trading
\item Recommending movies
\item Online search
\item Self driving cars
\end{itemize}

\subsubsection*{What aspects of a problem determine if causal inference is required?}
(When is pure prediction useful?)
\begin{itemize}
\item To decide between actions we only need to rank them (not estimate their actual effect). 
\item The predicted outcome in the absence of an intervention provides a single point. We can use this to find which problems are most serious if left alone - and priortise those for modelling changes. 
\item Any decision we take does not significanly impact the system from which the data was drawn to make it (for repeat decision making)
\item Does acting on the result of the prediction change the predictive distribution p(y|x)? Ie change people's behaviour.

\end{itemize}

@@

@@

@@

\chapter*{Causal models (5000 words)}
\subsection*{Causal graphical models}
Although seemingly simplistic, the notion of hard interventions is surprisingly powerful.

@@

\subsection*{Counterfactuals}
\subsection*{Structural Equation models}
\subsection*{Comparing and unifying the models}
\subsubsection{A translator from graphical independence to counter factual statements}

\chapter*{Two key questions (5000 words)}
\section*{Causal Inference} 
\paragraph*{You are willing to assume the causal graph}
\paragraph*{extremely widely appliled} Implicitly accounts 10,000 of studies in phycology, medicine, business, etc.

\subsection*{Identifiability}
\paragraph*{Under what conditions in is the problem solveable}
\subsubsection*{The Do Calculus}

\subsubsection*{Open questions within identifiablity}

\subsection*{Estimation}
\paragraph*{How well do we actually do with finite datasets?}

\section*{Causal Discovery}
\paragraph*{You want to learn the graph}
\paragraph*{Equates to the aim of automating scientific discovery}
\paragraph*{Increadibly hard}
\subsection*{Discovery with Conditional independence}
\subsection*{Discovery with Functional Models}

\chapter*{Multi-Armed Bandits (5000 words)}

>>>>

\documentclass[11pt,a4paper,oneside]{book}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{dsfont}
\usepackage{listings}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{tabularx}
\usepackage{enumitem}
\usepackage{algorithm}% http://ctan.org/pkg/algorithm
\usepackage[noend]{algorithmic}
\usepackage{parskip}
\usepackage{tikz}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{epigraph}
\usetikzlibrary{arrows,positioning} 
\usepackage{subcaption}
\usepackage[colorinlistoftodos]{todonotes}
\usepackage{natbib}
\usepackage{wrapfig}

\graphicspath{ {figures/} }
\pgfarrowsdeclarecombine{ring}{ring}{}{}{o}{o}

@@

\newcommand{\KL}{\operatorname{KL}}
\newcommand{\simpleregret}{R_T}
\newcommand{\Pij}[1]{\operatorname{P}_{ij}\!\left\{#1\right\}}
\newcommand{\Pkl}[1]{\operatorname{P}_{kl}\!\left\{#1\right\}}
\newcommand{\Q}[1]{\operatorname{Q}\left\{#1\right\}}
\newcommand{\EE}{\mathbb E}
\newcommand{\EEa}{\EE_a}
\newcommand{\Pns}[2]{\operatorname{P}_{#1}\left\{#2\right\}}
\newcommand{\Pn}[2]{\operatorname{P}\left\{#2|#1\right\}}
\newcommand{\actions}{\mathcal{A}}
\newcommand{\calA}{\mathcal A}
\newcommand{\etc}{\textit{etc}}
\newcommand{\ie}{\textit{i.e.}}
\newcommand{\eg}{\textit{e.g.}}
\newcommand{\calP}{\mathcal P}
\newcommand{\x}{\boldsymbol{x}}
\newcommand{\Ps}{\operatorname{P}}
\newcommand{\R}{\mathbb R}
\newcommand{\Pri}[1]{\operatorname{P}_i\left\{#1\right\}}
\newcommand{\Prz}[1]{\operatorname{P}_0\left\{#1\right\}}

@@

\renewcommand{\P}[1]{\operatorname{P}\left\{#1\right\}}
\newcommand{\cf}[2]{{#1}_{#2}}
\newcommand{\bernoulli}{\operatorname{Bernoulli}}
\newcommand{\dirac}{\operatorname{Dirac}}
\newcommand{\parents}[1]{\operatorname{\mathcal{P}a}_{#1}}
\renewcommand{\vec}[1]{\boldsymbol{#1}}

%\theoremstyle{plain}

\newtheorem{theorem}{Theorem}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}

%\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{assumption}[theorem]{Assumption}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{example}[theorem]{Example}
\let\temp\epsilon
\let\epsilon\varepsilon

@@

\begin{document}
\todo{figure out theoremstle is undefined}
\def\ci{\perp\!\!\!\perp} % from Wikipedia

\maketitle

\epigraph{In vain the grave, with retrospective Eye,
Would from the apparent what conclude the why,
Infer the Motive from the Deed, and show
That what we chanced, was what we meant, to do.}{Alexander Pope}

3100 words at the start of boot camp.

\chapter*{Random notes}

Bandit feedback. The learner observes only the reward for the action they select. Not every possible option as in supervised learning.

@@

\paragraph*{My thesis in a sentence:} Unifying causal inference with multi-armed bandits.

\paragraph*{This thesis contributes to knoweldge by:} Introducing a framework connecting causal graphical models with multi-armed bandits as a first step towards a unified approach to estimating the effect of interventions.

\paragraph*{My key research questions are:} 
\begin{itemize}
\item To understand and the difference between prediction and causal inference in machine learning and clarify which problems require causal approaches.
\item To summarize the key strands of causal inference research from economics and social sciences for the machine learning community
\item To make connections between learning to act from observational versus experimental data. In particular, between causal graphical models and multi-armed bandits.
\end{itemize}

\section*{Motivation}
Many of the most important questions in science and in our personal lives are about the outcomes of doing something. Will asking people to pay upfront at the doctors reduce long term health expenditure? If we developed a drug to suppress particular genes, could we cure MS and would delaying teen-aged pregnancies improve the outcome for their kids.

These are hard questions because they require more than identifying a pattern in data. Correlation is not causation. Causal inference has proven so difficult that there is barely any consensus on even enduring questions like the returns to education or the long-term consequences of early life events – like teenage pregnancy, where the variables involved are susceptible to human intuition and understanding.

We now live in a world of data. Hours of our lives are spent online, where every click can be recorded, tiny computers and sensors are cheap enough to incorporate into everything and the US Institute of Health is considering if all infants should be genetically sequenced at birth. Such data gives us a window into many aspects of our lives at an unprecedented scale and detail but it is messy, complicated and often generated as a by product of some other purpose. It does not come from the controlled world of a randomised experiment.

The rise of big data sets and powerful computers has seen an explosion in the application of machine learning. From healthcare, to entertainment and self driving cars; machine learning algorithms will transform many industries. It has been suggested that the impressive ability of statistical machine learning to detect complex patterns in huge datasets heralds the end of theory (Reference) and that we may be only a short step from the Singularity, where artifcial intelligence exceeds our own and then grows exponentially.

However, despite the huge advances in specific areas of machine learning (in particular deep learning), machine learning algorithms are effective only within very narrow problem settings. Getting them to generalize to even slightly different problems or datasets remains very challenging. Deciding how we should act or what policies we should implement requires us to be able to predict how a system will behave if we change it. The correlations detected by standard machine learning algorithms do not enable us to do this, no matter how many petabytes of data they are based on. As machine learning algorithms are incorporated into more and more of the decision making processes that shape the world we live it, it is critical to ensure that we understand the distinction between causality and prediction and that we develop techniques for learning how to act that are as effective as those we have for pattern recognition.

@@

To explain or to predict \cite{Shmueli2010}
The two cultures \cite{Breiman2001}

\subsection*{Defining causality}
\begin{itemize}
\item widely debated in science and philosophy (REFERENCES)
\item what is explaination?
\item any model that aims to predict the outcome of an action or intervention in a system
\item I do not see the distinction between explaination and (causal) prediction. Explaination is all about the ability to compress and to generalize. The more a model can do this, the more we view as providing an understanding of the why. 
\item mediation?
\end{itemize}

@@

\begin{itemize}
\item Speech recognition (for systems like Siri or Google)
\item Machine translation 
\item Image classification
\item Forecasting the weather
\item Playing Go 
\item Identiying spam emails
\item Automated essay marking
\item Predicting the risk of death in patients with pnumonia.
\item Predicting who will re-offend on release from prison 
\item Predicting which customers will cease to be your customers
\item Demand prediction for inventory control
\item Predicting who will click on an ad
\item Financial trading
\item Recommending movies
\item Online search
\item Self driving cars
\item Pricing insurance
\end{itemize}

%For image recognition, we do not particularly care about building a strong model for exactly how the thing that was photographed translates to the image we see. In fact this is because we already have one. We can be confident that the process will not change. If we develop a discriminative model that is hightly accurate at classifying cats from dogs, we do not care a lot about its internal workings (provided we have strong grounds to believe that the situations in which we will be using our model will match those under which it was trained). Covariate shift clearly comes in here. Because there are areas where mechanisms are understood it is relatively easy to argue that covariate shift is not occuring and that results will be transferable. The mechanism is known but the function may be complex.

\subsubsection*{What aspects of a problem determine if causal inference is required?}
(When is pure prediction useful?)
\begin{itemize}
\item To decide between actions we only need to rank them (not estimate their actual effect). 
\item The predicted outcome in the absence of an intervention provides a single point. We can use this to find which problems are most serious if left alone - and priortise those for modelling changes. 
\item Any decision we take does not significanly impact the system from which the data was drawn to make it (for repeat decision making)
\item Does acting on the result of the prediction change the predictive distribution p(y|x)? Ie change people's behaviour.
\item Ethics - ... People's viewpoint on if its ok...
\end{itemize}

@@

There are two broad approaches to deciding how to act. Reinforcement learning and causal inference. In reinforcement learning we estimate the effect of actions by taking them. We assume there is an agent capable of intervening in the system and try to find policies for the agent to follow in selecting actions so as to maximise some kind of reward. This is a very powerful and general framework (REFERENCES TO GENERAL AI). However, POINT OUT SOME OF THE DIFFICULTIES WITH REINFORCEMENT LEARNING. We are frequently presented with large bodies of data that have been collected from a system in which we did not have any control over what actions were taken.

We will call datasets where we do not have control over the decision making process that generated the data observational. Versus experimental datasets, where we do have control (experimental datasets do not always have to be randomized -although that is a powerful approach to ensure we have control. There is a space in the middle where we have partially controlled the process by which agent select actions. Randomized data with imperfect compliance would be an example.

An agent (capable of intervening in the system) chooses an action from those availble The agent making the decision included in the model. This is

Reinforcement learning addresses the problem of learning from explicitly taking actions. There
is typically some state or environment. An agent chooses an action from those available in the
current state. The state then evolves stochastically as a function of the selected action and
the agent receives some feedback or reward that is a function of the new state. This setting
differs from the standard classification problem in that the agent must learn from feedback on
the selected action, rather than being presented with the correct action for a given state. A
common modelling assumption is that the state evolves only as a function of the previous state
and the action chosen, and given these, is independent of the previous history of states and
actions. This is known as a Markov decision process or MDP. A particularly well studied and
understood model is the singe state MDP. In this case, there are a set of actions, each associated
with a fixed but unknown reward distribution and at each time step our agent selects an action
and receives corresponding feedback. This is known as the multi-armed bandit problem.

Causal inference makes use of assumptions to allow the outcome of actions to be predicted from
observational data. The key to causal inference is a framework that can model how actions
change the state of the world. This framework then allows us to map information collected in
one setting to another.

Both approaches can be seen as extensions to the concept of randomised controlled trials. Bandit
algorithms deal with the sequential nature of the decision making process, causal inference with
the problem that full randomisation is not always feasible, affordable or ethical. The similarities
between the problems that these techniques have been developed to address raises the question
of if there are problems best addressed by a combination of these approaches and how they
can be combined. The goal of my thesis is to explore these questions. In the next sections I
review the key literature in causal inference and bandits. I then present a general approach to
how causal models might be incorporated into bandit settings and conclude by demonstrating
an algorithm that leverages causal assumptions to improve performance in a specific bandit
setting.

There are two key approaches to causal problems. The first is to learn the outcome of actions by directly intervening in the system and seeing what happens. We then get feedback on how good those actions were. This is the approach taken in reinforcement learning. ADVANTAGES AND DISADVANTAGES OF THIS APPROACH. The second broand approach is causal inference. Here

@@

Both these fields relate to the problem of making optimal decisions and both can be seen as generalising randomised controlled experiments. Causal inference  is the study of how to estimate the effect of an action in the absence of randomisation. Reinforcement learning studies how we can do better if the decisions are to be made sequentially.

@@

\chapter*{Causal models (3000 words)}

Causal inference aims to infer the outcome of an intervention in a system from data obtained by observing (but not intervening in) the system. To do this we need terminology to describe actions and how we anticipate the system should respond to them. Three key approaches have emerged; counterfactuals, structural equation models and causal bayesian networks. In this chapter we describe these approaches. We will examine what problems they allow us to solve, what assumptions they rely on and how they differ. We will also use them to describe the following very simple example. The aim is to demonstrate the notations and formalisms we will need to tackle more interesting problems later on.

Suppose we have developed a new drug for some illness and wish to determine how effective it is. We take a large group of patients and randomly assign half of them to a treatment group and the other half to a control group. The people in the treatment group get the drug, everyone else gets a placebo pill. The question we want to answer is does giving people the active drug improve their changes of recovery relative to giving them the placebo. We will use the variable $X$ (1 = drug, 0 = placebo) to represent the treatment each person receives and $Y$ (1 = recover, 0 = not recover) to describe the outcome.

\subsection*{Causal bayesian networks}

Causal Bayesian Networks are an extension of Bayesian networks. A bayesian network is a graphical way of representing how a distriution factorises. Any joint probability distribution can be factorized into a product of conditional probabilities. There are multiple valid factorizations, corresponding to permutations of variable ordering.

\begin{equation}
P(X_{1},X_{2},X_{3},...)=P(X_{1})P(X_{2}|X_{1})P(X_{3}|X_{1},X_{2})...
\end{equation}

We can represent this graphically by drawing a network with a node for each variable and adding links from the variables on the right hand side to the variable on the left for each conditional probability distribution (figure~\ref{fig:bayesnet}). If there are conditional independencies between variables the factorization simplifies, which is reflected by missing edges in the corresponding network. 
%
%\begin{figure}[h]
%\centering
%\caption{General Bayesian network over 3 variables}
%\label{fig:bayesnet}
%\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
%  thick,main node/]
%
% %nodes
%\node[main node](1){$X_{1}$};
%\node[main node, below left=of 1](2){$X_{2}$};
%\node[main node, below right=of 1](3){$X_{3}$};
%
%
% \path[every node/.style={font=\sffamily\small}]
%    (1) edge node {} (2)
%    	edge node {} (3)
%    (2) edge node {} (3);
%	
%\end{tikzpicture}
%\end{figure}

The statement that a given graph $G$ is a Bayesian network for a distribution $P$ tells us that the distribution can be factorized over the nodes and edges in the graph. There can be no missing edges in $G$ that do not correspond to conditional independencies in $P$ (the converse is not true $G$ can have extra edges). If we let $parents_{X_{i}}$ represent the set of variables that are parents of the variable $X_{i}$ in $G$ then we can write the joint distribution as;

\begin{equation}
P(X_{1}...X_{N}) = \prod_{i = 1...N}P(X_{i}|parents_{X_{i}})
\end{equation}

A causal Bayesian network is a Bayesian network in which it a link $X_{i} \rightarrow X_{j}$, by definition, implies $X_{i}$ causes $X_{j}$. This means that if we intervene and change the value of $X_{i}$, we expect $X_{j}$ to change, but if we intervene to change $X_{j}$, $X_{i}$ will not change. We need some notation to represent the distribution after an intervention. In this thesis we will use the do opperator introduced by Pearl \ref{XXX}, where $\P{Y|do(X=x)}$ represents the distribution of the variable $Y$ after an intervention that sets the variable(s) $X$ to $x$. TALK ABOUT HOW THIS CAN REPRESENT MULTIPLE THINGS CONTINUOUS VERSUS DISCRETE DISTRBUTIONS AND WHETHER OR NOT VARIABLES ARE RESOLVED OR AS A FUNCTION. It is important to understand that in general, the interventional distribution $P{Y|do(X=x)}$ will differ from the standard conditional distribution $\P{Y|X=x}$.

If $G$ is a causal network for a distribution $P$ defined over variables $X_{1}...X_{N}$, then we can calculate the distribution after an intervention where we set $Z \subset X$ to $z$, denoted $do(Z=z)$ by simply dropping the terms for each of the variables in $Z$ from the factorization given by the network. This is referred to as the truncated product formula \cite{Pearl2000}.

\begin{equation}
\label{eq:truncatedproduct}
 P_{do(Z=z)}(X_{1}...X_{N}) =
  \begin{cases}
  \prod_{i \notin Z}P(X_{i}|parents_{X_{i}}) & \text{if $(X_{1}...X_{N})$ consistant with $Z=z$}  \\
   0       & \text{otherwise } 
  \end{cases}
\end{equation}

This result does not hold for standard bayesian networks for the simple reason that there are multiple valid networks for the same distribution. The truncated product formula will give different results depending on which you choose. The result is possible with causal bayesian networks because it follows directly from the assumption that the direction of the link indicates causality. In fact, from the interventionalist viewpoint causality the truncation product formula defines what it means for a link to be causal.

Returning to our simple example, and phrasing our query in terms of interventions; what would the distribution of outcomes look like if everyone was treated $P(Y|do(X=1))$, relative to if no one was treated $P(Y|do(X=0))$? Since the treatment $X$ is assigned  via deliberate randomisation we know there it is not effected by any unobserved variables. $X$ is a potential cause of $Y$, along with other unobserved variables, such as the age, gender and desease subtype of the patient.

The causal bayesian network for this scenario is:

\todo{proper picture}
\eq{
X \rightarrow Y\\
U_l \rightarrow Y
}

This network represents the (causal) factorization  $P(X,Y) = P(X)P(Y|X)$, so from equation (\ref{eq:truncatedproduct}), $P(Y|do(X)) = P(Y|X)$. In this example, the interventional distribution is equivelent to the observational one.

Formalising the definition of an intervention within the framework of causal graphical models provides us with an explicit mechinism to map information from one data generating process, the system pre-intervention, to another, the system post-intervention. A key feature of this way of defining an intervention is the number of things that are invariant to the intervention. That is, all the conditional distributions over variables that were not intervened are assumed not to be changed by the intervention.

\todo {draw the system as two seperate models}

A causal bayesian network encodes all possible interventions that can be represented with the do notation. If you knew of all the terms in the causal factorisation, then you could compute the distribution over any variable in the graph after setting any subset of other vairables (subject to possitivity assumption) 
\todo {state positivity assumption}

A causal network represents much more information than a bayesian network with identical structure. GIVE EXAMPLE IN SOMETHING QUANTIFIABLE LIKE BITS. This may be surprising since their conditional probability distributions are estiated from data in the same way. It is only possible due to the additional assumptions inherent in a causal network.

Causal bayesian networks are bayesian networks so results that apply to bayesian networks carry directly across;the local markov property states that variables are independent of their non-effects given their direct causes. Similarly the global markov property and d-separation also hold in causal networks.

\subsubsection*{Limitations of causal bayesian networks}
A number of critisims have been leveled at this way of modelling causality.

@@

What questions can't be phrased using this terminology
- certain counterfactual queries
- cyclic/dynamics

Although seemingly simplistic, the notion of hard interventions is surprisingly powerful.

- add variables for soft interventions
- unroll to deal with cycles

\subsection*{Structural Equation models}

There is a problem with the term non-parametric. It is used in machine learning in a slightly different way to pearl

\subsubsection*{Definition}
Structural equation models (SEMs) describe a deterministic world, where some underlying mechanism or function determines the output of any process for a given input. The mechanism (but not the output) is assumed to be independent of what is fed into it. Uncertainties are not inherent but arise from unmeasured variables. Linear structural equation models have a long history for causal estimation \cite {Wright1921,Haavelmo1943}. More recently, they have been formalized, generalized to the non-linear setting and connected to developments in graphical models to provide a powerful causal framework \cite{Pearl2000}.

Mathematically, each variable is a deterministic function of its direct causes and a noise term that captures unmeasured variables. The noise terms are required to be mutually independent. If there is the possibility that an unmeasured variable influences more than one variable of interest in a study, it must be modelled explicitly as a latent (unobserved) variable. Structural equation models can be represented visually as a network. Each variable is a node and arrows are drawn from causes to their effects. For our example the SEM is:

This model encodes the assumption that the outcome $y_{i}$ for an individual $i$ is caused solely by the treatment $x_{i}$ they receive and other factors $\epsilon_{y_{i}}$ that are independent of $X$. This is justifiable on the grounds that $X$ is random. The outcome of a coin flip for each patient should not be related to any of their characteristics (hidden or otherwise).

\subsubsection*{Example problem}
We want to estimate the causal effect of treatment; what is the probability of recovery if we \textbf{take the action} 'treat' versus the \textbf{action} 'placebo'? Taking the action 'treat'  corresponds to replacing the equation $X = \epsilon_{x}$ with $X=1$. The probability distribution over $Y$ given we set $X=1$ is then $P(Y = y|do(X=1)) = P(f(1,\epsilon_{y})=y)$, where we have introduced the $do$ notation to distinguish setting a variable from observing it \cite{Pearl1995}. However, for this model,  The probability of observing $Y$ given $X=1$, $P(Y = y|X = 1)$ is also given by $P(f(1,\epsilon_{y})=y)$ because $\epsilon_{y} \ci \epsilon{x}$. The causal effect is exactly the difference in observed outcomes, as we would intuitively expect for a randomized experiment. In this case, due to the assumption that $X \rightarrow Y$ and that there is no hidden common cause, correlation is causation.

For a model with $N$ variables, a structural equation model looks like a set of $N$ simultaneous equations, with each variable playing the role of the dependent (left hand side) variable in one equation. However a SEM is, by definition, more than a set of simultaneous equations. By declaring it to be structural we are saying that it represents assumptions about the relationships between variables. When we visualise the model as a network the absence of an arrow between two variables encodes the assumption that one does not cause the other.

Limitations \& Critisisms 
- can express some non-measurable things
	- example of that

- confusion may occur as to whether interpretation is causal

- can they handle cycles?

\subsection*{Counterfactuals}
\todo{use cf command to in counterfactual equations}

The Neyman-Rubin model \cite{Rubin1974,Rubin1978,Rosenbaum1983, Rubin2005,Rubin2008} defines causality in terms of potential outcomes, or counterfactuals. Counterfactuals are statements about imagined or alternate realities, are prevalent in everyday language and may play a role in the development of causal reasoning in humans \cite{Weisberg2013}. Causal effects are differences in counterfactual variables; what is the difference between what would happen if we did one thing versus what would happen if we did something else.

In our example, the causal effect of the drug relative to placebo for person $i$ is the difference between what would happen if they were given the drug, denoted $y_{i}^{1}$ versus what would happen if they got the placebo, $y_{i}^{0}$. The fundamental problem of causal inference is that we can only observe one of these two outcomes, since a given person can only be treated or not treated. The problem can be resolved if, instead of people, you have units you can assume are identical or that will revert exactly to their initial state some time after treatment. This type of assumption often holds to a good approximation in the natural sciences and explains why researchers in these fields are less concerned with causal theory.

Instead of trying to estimate individual effects, lets see if we can learn something about the distributions under treatment or placebo.  Let $\cf{Y}{1}$ be a random variable representing the potential outcome if treated. The distribution of $\cf{Y}{1}$ is the distribution we would see of $Y$ if everyone was treated. Similarly $Y^{0}$ represents the potential outcome for the placebo. We want to know the difference between the probability of recovery, across the population if everyone was treated, and the probability of recovery given placebo  $\P{\cf{Y}{1}}-\P{\cf{Y}{0}}$. We can estimate, from an experimental or observational study, the probability that people recover if treated $P(Y|X=1)$ and the probability that they recover if not treated $P(Y|X=0)$. Now if $X=0$ then $Y = \cf{Y}{0}$. Equivalently stated:

\eq{
\P{Y^{0}|X=0}&= \P{Y|X=0}\\
\P{Y^{1}|X=1}&=P{Y|X=1}
}

If we assume $X \ci Y^{0}$ and $X \ci Y^{1}$:

\eq{
\P{Y^{1}} &= \P{Y^{1}|X=1} = \P{Y|X=1} \\
\P{Y^{0}} &= \P{Y^{0}|X=0} = \P{Y|X=0}
}

\eq{
\implies \P{Y^{1}}-\P{Y^{0}} = \P{Y|X=1} - \P{Y|X=0}
}

The assumptions $X \ci Y^{1}$ and $X \ci Y^{0}$  are referred to as ignoreability assumptions \cite{Rosenbaum1983}. They state that the treatment a each person receives is independent of whether they would recover if treated and if they would recover if not treated. Again this is justified in our example due to the randomization of treatment assignment. In general these assumption do not hold. If people were deciding whether or not to buy the treatment, rather than it being randomly assigned, there could be a variable, for example income,D-separation still applies in the augmented model. that influenced both the decision to get treatment and the likelyhood of recovery given treatment or placebo.

\todo{SUTVA assumption}

There are some complex philosopical objections to counter factuals arising from the way they describe alternate universes that were never realized and are not empirically testable \ref{XXXXX}. This can result in some concrete problems. Consider the following example based on  \ref{Dawid2000}.

Again we have a drug, where the outcome for an individual if treated is represented by the counterfactual variable $\cf{Y}{1}$ and the outcome if not treated is $\cf{Y}{0}$. Suppose these counterfactual variables $\cf{Y}{1}$ and $\cf{Y}{0}$ are jointly normal. We will assume equal variance for simplicity.

% vector with mean mu_t,mu_c and a covariance matrix
\eq{
P(Y_t,Y_c) \sim N()
}

then their difference is also normal,

\eq{
P(\tau) = N(\mu_t - \mu_c,2\sigma^{2}(1-\rho))
}

\includegraphics[scale=.6]{figures/counterfactual_nonidentify.pdf}

Key issue is that we can never observe the joint distribution over $Y_t$ and $Y_c$. As a result, the variance of $\tau$ is not identifiable, even from experimental data. It seems on the face of it that $\tau$ is relevent to our decision making. If the example above, if $\rho = 1$ then almost everyone benifits slightly from the treatment. However if $\rho=0$, there is a wide range, with some people benifiting a lot and others suffering significant harm.

Can these issues be resolved? If so, how?

- note that we can bound this counterfactual distribution based on the variance of the observed interventional distributions. If the variance of XXX is small then this may not be an issue.

- does it actually make sense to make a decison on the basis of this counterfactual that is a function of something we could never observe. I

- part of the problem may be due to the deteminisitic way we have phrased individual treatment effects.

Can these issues be resolved by considering personalized causal effects as random variables and avoiding counterfactuals?

A range of similar issues can arise in counterfactuals \ref{XXXX}. 
- What assumptions are required for the kind of counterfactual analysis like X would have been higher had B ...?
- briefly discuss the different flavours of counterfactual questions here.

One way of looking at counterfactuals is as a natural language short hand for describing highly specific interventions like those denoted by the do-notation. Rather than taking about the distribution of Y given we intervene to set $X=x$ and hold everything else about the system constant we just say what would the distribution of $Y$ be had $X$ been $X$. This is certainly convienient, if rather impressise.

\subsection*{Comparing and unifying the models}

Remarkably for models developed relatively independently in fields with very different approaches and problems, the models we have discussed can be nicely unified for interventional queries (those that can be expressed with the do-notation). This makes it straighforward to combine key results and algorithms developed within any of these frameworks. For example, draw a graphical network to determine if a problem is identifiable and which variables should be adjusted for to obtain an unbiased causal estimate. Then use propensity scores \ref{} estimate the effect. If non-parametric assumptions are insufficient for identification or lead to overly large uncertainties, you can specify additional assumptions by phrasing your model in terms of structural equations.

If the network for a structural equation model is acyclic, that is if starting from any node and following edges in the direction of the arrows you cannot return to the starting point, then it implies a recursive factorization of the joint distribution over its variables. In other words, the network is a causal Bayesian network. All of the results that apply to causal Bayesian networks also apply to acyclic structural equation models.  Taking an action that sets a variable to a specific value equates to replacing the equation for that variable with a constant. This corresponds to dropping a term in the factorization and the truncated product formula (equation \ref{eq:truncatedproduct}). Thus, the interventional query $P(Y|do(X))$ is identical in these two frameworks. We can also connect this to counterfactuals via:

\begin{equation}
\begin{aligned}
&Y^{0} \equiv P(Y|do(X=0)) \\
&Y^{1} \equiv P(Y|do(X=1))
\end{aligned}
\end{equation}

The assumption $\epsilon_{X} \ci \epsilon_{Y}$, stated for our structural equation model, translates to $X \ci (Y^{0},Y^{1})$ in the language of counterfactuals. When discussing the counterfactual model, we actually made the slightly weaker assumption:

\begin{equation}
\label{eq:weakignore}
X \ci Y^{0} \text{ and } X \ci Y^{1}
\end{equation}

It is possible to relax the independence of errors assumption we made for SEMs slightly to correspond exactly the form of equation (\ref{eq:weakignore}) without losing any of the power provided by d-separation and graphical identification rules \cite{Richardson2013}. To determine if and how an interventional query can be non-parametrically identified, it is equivalent to specify assumptions graphically in terms of functional models or bayesian networks or as conditional independence statements involving counterfactual variables (ignorability assumptions). By non-parametrically, I mean that we are not making any assumptions about the form of the relationships between variables. Models that are not non-parametrically identifiable can still be identified given assumptions about the distributions of variables and the functional relationship between them, for example, that the functions are linear or that the noise is additive \cite{Peters2014}. This form of assumption fits extremely naturally into the structural equation framework.

However we can also pose causal queries that are not interventional and cannot be phrased in terms of the do-notation. The patients in our drug treatment example could be broken down into four groups. \todo{point out objections to doing this}. The first group will recover whether or not they receive treatment, the second group will recover if treated but not on the placebo, the third group will recover on the placebo and not if treated, and the last group will not recover on treatment or placebo. Unfortunately, we don't know which group each person belongs to. Drawing this up as a table:

\begin{tabular}{c|c|c|c}
group & placebo & treatment & probability of group\\
\hline
1 & die & die & $\alpha=P(Y^{0}=0,Y^{1}=0)$\\
2 & die & recover & $\beta=P(Y^{0}=0,Y^{1}=1)$\\
3 & recover & die & $\gamma=P(Y^{0}=1,Y^{1}=0)$\\
4 & recover & recover & $\delta=P(Y^{0}=1,Y^{1}=1)$\\
\end{tabular}

The queries we have been asking thus far are about $P(Y^{0}=1) = \gamma + \delta$ and $P(Y^{1}=1) = \beta + \delta$, but suppose we asked the question; what is the probability that this patient, who was not treated and died, would have recovered if they had been treated? We know they are in either group 1 or 2 since they died without treatment, so the answer is $\frac{\beta}{\alpha+\beta}$. Can we estimate the $\alpha, \beta, \gamma, \delta$ or in other words, identify the joint distribution over the counterfactuals $P(Y^{0},Y^{1})$ given the interventional distributions, $P(Y^{0})$ and $P(Y^{1})$? The answer is no, putting our constraints and unknowns in matrix form:

\begin{equation}
\left(
\begin{array}{cccc}
0&0&1&1\\
1&1&0&0\\
0&1&0&1\\
1&0&1&0\\
1&1&1&1\\
\end{array}
\right)
\left(
\begin{array}{c}
\alpha\\
\beta\\
\gamma\\
\delta\\
\end{array}
\right)= 
\left(
\begin{array}{c}
P(Y^{0})\\
1-P(Y^{0})\\
P(Y^{1})\\
1-P(Y^{1})\\
1\\
\end{array}
\right)
\implies
\left(
\begin{array}{c}
\alpha\\
\beta\\
\gamma\\
\delta\\
\end{array}
\right)=
\left(
\begin{array}{c}
1-P(Y^{0})-P(Y^{1})+\delta\\
P(Y^{1})-\delta\\
P(Y^{0})-\delta\\
\delta\\
\end{array}
\right)
\end{equation}
The value of $\delta$ is not determined so the query is not identifiable. However we do get bounds on the terms. Since probabilities cannot be negative, $P(Y{1})-P(Y^{0})-1 \leq \delta \leq min(P(Y^{1}),P(Y^{0})$. Note, if we made the additional assumption $\gamma=0$; that the drug did not cause anyone to die who would otherwise have survived, then we can determine the joint distribution over counterfactuals. Alternatively, if we could assume that after treatment people returned to their initial state after some period of time, (say we were testing a drug for acne) then we could run a crossover study to determine the joint distribution. In a crossover study, the participants are randomly assigned to treatment and placebo, results are measured and then the groups are swapped. The scientific and philosophical validity of counterfactual queries remains under question \cite{Dawid2000,Dawid2014}, however they are nonetheless widely posed in the form of attribution of causal effects to different pathways and mediation \cite{Pearl2014,Imai2010a,VanderWeele2011}.
